---
title: "Article: Just enough ggplot"
author: Ben Stenhaug
output: 
  html_document:
    toc: true
    toc_depth: 1
    toc_float: true
    theme: default
    highlight: default
---

# Getting started

```{r message = FALSE}
library(tidyverse)
theme_set(theme_bw(base_size = 12))
set.seed(1)
```

This website — https://ggplot2.tidyverse.org/reference/ — is incredibly useful. We'll focus on the most basic building blocks. Let's build a couple of data frames to use:

```{r}
scores <- 
  tibble(
    name = c("mike", "carol", "greg", "marcia", "peter", "jan", "bobby", "cindy", "alice"),
    school = c("south", "south", "south", "south", "north", "north", "north", "south", "south"),
    teacher = c("johnson", "johnson", "johnson", "johnson",  "smith", "smith", "smith", "perry", "perry"),
    math_score = c(4, 3, 2, 4, 3, 4, 5, 4, 5),
    reading_score = c(1, 5, 2, 4, 5, 4, 1, 5, 4)
  )

random_numbers <- 
    tibble(
        binom = rbinom(1000, 10, 0.5),
        pois = rpois(1000, 5)
    ) %>% 
    mutate(
        mix = rnorm(1000, binom + pois, 5),
        high_low = ifelse(mix > 10, "high", "low")
    )
```

To get the basics down, I'll show an example using the `scores` dataset and then ask you to make a similar plot using the `random_numbers` dataset.

# Plotting one variable

## geom_histogram

If your variable is continuous, you'll want to use `geom_histogram`:

```{r}
scores %>% 
    ggplot(aes(x = math_score)) +
    geom_histogram(binwidth = 2)

?geom_histogram
```

Make a histogram with the random numbers and play around with `geom_histogram` options:

```{r}
random_numbers %>% 
  ggplot(aes(x = binom)) +
  geom_histogram(bins = 20)
```

## geom_bar

If your variable is discrete, you'll want to use `geom_bar`:

```{r}
scores %>% 
    ggplot(aes(x = teacher)) +
    geom_bar()
```

Your turn with `geom_bar`:

```{r}
random_numbers %>% 
  ggplot(aes(x = high_low)) +
  geom_bar() +
  coord_flip()
```

# Plotting two variables 

## geom_point

It's most common to use a scatter plot with `geom_point`:

```{r}
scores %>% 
    ggplot(aes(x = math_score, y = reading_score)) +
    geom_point()
```

You don't have to use continuous variables!

```{r}
scores %>% 
    ggplot(aes(x = math_score, y = name)) +
    geom_point()
```

Your turn:

```{r}

```

## geom_col

Making a column for each "point" with `geom_col` can also be useful:

```{r}
scores %>% 
    ggplot(aes(x = name, y = reading_score)) +
    geom_col() +
    geom_point(color = "red")
```

Your turn:

```{r}

```

## geom_boxplot

`geom_boxplot` is useful for plotting a discrete continuous variable:

```{r}
scores %>% 
    ggplot(aes(x = school, y = math_score)) +
    geom_boxplot()
```

Your turn:

```{r}

```

## geom_smooth

`geom_smooth` is a fancy geom but still just uses two variables:

```{r}
scores %>% ggplot(aes(x = reading_score, y = math_score)) + geom_smooth(method = "lm", se = FALSE)  + geom_point()
```

You can have multiple geoms — in this case it's common to point points and a smoother both on a plot.

Your turn:

```{r}

```

# Adding a third variable

There are a few ways:

```{r}
# color 
scores %>% 
    ggplot(aes(x = reading_score, y = math_score, color = school)) +
    geom_point()

# shape
scores %>% 
    ggplot(aes(x = reading_score, y = math_score, shape = school)) +
    geom_point()

# with a facet
scores %>% 
    ggplot(aes(x = reading_score, y = math_score)) +
    geom_point() +
    facet_wrap(~ school)
```

Your turn:

```{r}

```

# Putting it all together

Here's a crazy plot that uses many, many options that might be useful for reference:

```{r}
scores %>% 
    ggplot(mapping = 
             aes(x = math_score, y = reading_score, color = school)) +
    geom_point(alpha = 0.4) +
    scale_x_continuous(limits = c(0, 10)) +
    scale_y_log10() +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~ school, ncol = 1) +
    labs(
        x = "Math",
        y = "Reading",
        title = "Math and reading test scores by student",
        subtitle = "Just for 8 students"
    ) +
    theme_gray(base_size = 18)

# this is useful for a session: theme_set(theme_bw(base_size = 12))
```

Your turn to make a crazy plot:

```{r}
# go for it
```

