---
title: "Exercise: Speed"
author: Ben Stenhaug
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: default
    highlight: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# install.packages("tictoc")
library(tictoc)
```

Create data

```{r}
n_cols <- 100000

df <- 
    rerun(n_cols, rnorm(10)) %>% 
    as_tibble(.name_repair = "universal")
```

An example of telling time

```{r}
tic()
median(df$`1`)
toc()
```

# Many ways to calculate the median of each column

For each way, describe what the code is doing, grade the speed, and grade the human readability.

## Way 1

What's the code doing?

Speed:

Human readability:

```{r}
tic()
output1 <- vector("double", 0)
for (i in seq_along(df)) {
    output1 <- c(output1, median(df[[i]]))
}
toc()
```

## Way 2

What's the code doing?

Speed:

Human readability:

```{r}
tic()
output2 <- vector("double", ncol(df))
for (i in seq_along(df)) {
    output2[[i]] <- median(df[[i]])
    }
toc()
```

## Way 3

What's the code doing?

Speed:

Human readability:

```{r}
tic()
ouput3 <- apply(df, 2, median)
toc()
```

## Way 4

What's the code doing?

Speed:

Human readability:

```{r}
tic()
output4 <- lapply(df, median)
toc()
```

## Way 5

What's the code doing?

Speed:

Human readability:

```{r}
tic()
output5 <- df %>% map(median)
toc()
```

## Way 6

What's the code doing?

Speed:

Human readability:

```{r}
# install.packages("furrr")
library(furrr)
plan(multiprocess)

tic()
output6 <- df %>% future_map(median)
toc()
```

## Way 7

What's the code doing?

Speed:

Human readability:

```{r}
# install.packages("parallel")
library(parallel)
cl <- makeCluster(detectCores() - 1)

tic()
output7 <- 
    parLapply(
        cl, 
        df,
        median
    )
toc()
```

## Way 8

What's the code doing?

Speed:

Human readability:

```{r}
tic()
output8 <- df %>% summarize_all(median)
toc()
```

# Summarize

What are the factors that impact how fast R code is?

What are the factors that impact how human readable R code is?

# Learning more

- https://r4ds.had.co.nz/iteration.html

- https://www.r-bloggers.com/how-to-go-parallel-in-r-basics-tips/
